---
title: 'DS 2020 - Project Proposal'
author: "Eitan Tuchin, Tirmidi Mohamed"
date: "03/27/2025"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Proposal

### Research Topic

The topic for our research project is about bad-drivers in the US.

### Team Members

The team members working on this project are Eitan Tuchin and Tirmidi Mohamed.

### Description of Dataset

```{r}
data <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/bad-drivers/bad-drivers.csv")
head(data)
summary(data)
```
Our dataset contains varying statistics relating to specifically bad-drivers throughout the United States including the District of Columbia. For each state (row) there are 7 columns that contain statistics like percentage of drivers involved in fatal collisions who were impaired by alcohol use. Each metric is numerical, either being an int or a float. Most deal with the "number of" or "percentage of" drivers in that state.

### First Data Cleaning Steps

```{r}

```

### Marginal Summaries

### Project Idea (Questions we Are Answering)