---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis on Fatal Motor Vehicle Accidents in the United States
#### Eitan Tuchin, Tirmidi Mohamed

## Introduction
The goal of our project is to explore data about bad drivers in the United States so that we can better understand causes of fatal car crashes and point out common characteristics about bad drivers. Since driving is such a common activity that we as humans do on a day-to-day basis it becomes imperative to analyze times at when driving has gone wrong in the name of trying to reduce the risk for these dangerous situations. We hope that our exploration and the analysis of the data we looked at will lead to more effective driving policies throughout America and an altogether smaller fatalities reported in later data.

To achieve our project's goal we will answer the following questions:

  1. Can we identify regional patterns (e.g., Northeast, South, Midwest, West) in bad driving behaviors like speeding      and alcohol impairment, and how do these align with fatal collision rates? 

  2. Are there states where the percentage of speeding-related fatal collisions significantly deviates from the            national average, and what state-specific factors (e.g., speed limits, road conditions) might explain this? 

  3. How well can a predictive model (e.g., linear regression) use percentages of speeding, alcohol impairment, and        distraction to estimate fatal collisions per billion miles, and which factor is most influential?
  
These will be the main questions that we try and answer throughout our work on this project. We hope to get to some meaningful conclusions about bad drivers and their common trends in the United States.

## Data

## Results
```{r, warning=FALSE,message=FALSE}
library(readr)
library(tidyverse)
```

## Conclusion

