---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis on Bad Drivers in the United States
#### Eitan Tuchin, Tirmidi Mohamed

## Introduction
The goal of our project is to explore data about bad drivers in the United States so that we can better understand causes of fatal car crashes and point out common characteristics about bad drivers. Since driving is such a common activity that we as humans do on a day-to-day basis it becomes imperative to analyze times at when driving has gone wrong in the name of trying to reduce the risk for these dangerous situations. We hope that our exploration and the analysis of the data we looked at will lead to more effective driving policies throughout America and an altogether smaller fatalities reported in later data.

To achieve our project's goal we will answer the following questions:

  1. Can we identify regional patterns in bad driving behaviors like speeding and alcohol impairment, and how do          these align with fatal collision rates? 

  2. Are there states where the percentage of speeding-related fatal collisions significantly deviates from the           national average, and what state-specific factors might explain this? 

  3. How well can a predictive model use percentages of speeding, alcohol impairment, and distraction to                  estimate fatal collisions per billion miles, and which factor is most influential?
  
These will be the main questions that we try and answer throughout our work on this project. We hope to get to some meaningful conclusions about bad drivers and their common trends in the United States.

## Data
### Structure

The link to our dataset is https://github.com/fivethirtyeight/data/blob/master/bad-drivers/bad-drivers.csv. The github link points to a csv file under a repository. Our dataset contains varying statistics relating to specifically bad-drivers throughout the United States including the District of Columbia. For each state (row) there are 7 columns that contain statistics like percentage of drivers involved in fatal collisions who were impaired by alcohol use. Each metric is numerical, either being an int or a float. Most deal with the "number of" or "percentage of" drivers in that state.

Overall our dataset is pretty simple and it contains all of the data we need for answering our questions. 

### Cleaning

The first thing we need to do is read in the csv file and take a look at it:

```{r}
data <- read.csv("bad-drivers.csv")
summary(data)
head(data)
```

The first thing we did cleaning wise was check for missing values in our data by iterating over every column in our data and checking if the value was NA or not. Then we removed any columns with missing values.

```{r}
missing_values <- colSums(is.na(data))
print("Missing values per column:")
print(missing_values)
data <- data[, colSums(is.na(data)) == 0]  
print("Columns after removing those with NAs:")
print(names(data))
```

Then, for simplicity sake, we wanted to rename some of the columns in our data because they weren't as data explorable and accessible friendly. For example, we have a column named "Number of drivers involved in fatal collisions per billion miles" which would be annoying to access since it's separated by white-space and very long. We also checked to see if we got the intended result.

```{r}
colnames(data) <- c("State", "Fatal_Collisions_Per_Billion", "Percent_Speeding", 
                    "Percent_Alcohol", "Percent_Not_Distracted", "Percent_No_Prior", 
                    "Insurance_Premiums", "Losses_Per_Driver")
str(data)
```

Since we are going to be doing some statistical modeling later to answer our question we decided to turn the State column into a factor type from a String. Then we grabbed all the columns that contained numeric data and stored them. We then ensured that these specified columns are treated as numbers and not anything else. Now we have easy data to work with.

```{r}
data$State <- as.factor(data$State)
numeric_cols <- c("Fatal_Collisions_Per_Billion", "Percent_Speeding", "Percent_Alcohol", 
                  "Percent_Not_Distracted", "Percent_No_Prior", "Insurance_Premiums", 
                  "Losses_Per_Driver")
data[numeric_cols] <- lapply(data[numeric_cols], as.numeric)
```

Lastly, we need to check if we have any duplicate rows in our data. Then we removed those rows from the data.

```{r}
duplicates <- duplicated(data)
print("Number of duplicate rows:")
print(sum(duplicates))

data <- data[!duplicates, ]  

print("Final cleaned data:")
head(data)
```

After cleaning our data it's ready to be analyzed and interpreted.

### Variables

Here we list all of the variables (columns) in our dataset:

* State - The state to which the statistics in the row are attributed to.
* Fatal_Collisions_Per_Billion - Number of drivers involved in fatal collisions per billion miles.
* Percent_Speeding - Percentage of drivers who were involved in fatal collisions who were speeding.
* Percent_Alcohol - Percentage of drivers who were involved in fatal collisions who were alcohol-impaired.
* Percent_Not_Distracted - Percentage of drivers who were involved in fatal collisions who were not distracted.
* Percent_No_Prior - Percentage of drivers who were involved in fatal collisions who had not been in any previous accidents
* Insurance_Premiums - Average cost of car insurance premiums in that state.
* Losses_Per_Driver - Losses incurred by insurance companies for collisions per insured driver.

## Results

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(reshape2)
library(gridExtra)
```


### Can we identify regional patterns in bad driving behaviors like speeding and alcohol impairment, and how do these     align with fatal collision rates? 

```{r}
divisions <- list(
  `New England` = c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont"),
  `Middle Atlantic` = c("New Jersey", "New York", "Pennsylvania"),
  `East North Central` = c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin"),
  `West North Central` = c("Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska", "North Dakota", "South Dakota"),
  `South Atlantic` = c("Delaware", "Florida", "Georgia", "Maryland", "North Carolina", "South Carolina", 
                      "Virginia", "West Virginia", "District of Columbia"),
  `East South Central` = c("Alabama", "Kentucky", "Mississippi", "Tennessee"),
  `West South Central` = c("Arkansas", "Louisiana", "Oklahoma", "Texas"),
  `Mountain` = c("Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Utah", "Wyoming"),
  `Pacific` = c("Alaska", "California", "Hawaii", "Oregon", "Washington")
)

data <- data %>%
  mutate(Division = case_when(
    State %in% divisions$`New England` ~ "New England",
    State %in% divisions$`Middle Atlantic` ~ "Middle Atlantic",
    State %in% divisions$`East North Central` ~ "East North Central",
    State %in% divisions$`West North Central` ~ "West North Central",
    State %in% divisions$`South Atlantic` ~ "South Atlantic",
    State %in% divisions$`East South Central` ~ "East South Central",
    State %in% divisions$`West South Central` ~ "West South Central",
    State %in% divisions$`Mountain` ~ "Mountain",
    State %in% divisions$`Pacific` ~ "Pacific",
    TRUE ~ NA_character_
  ))


speed_alcohol_collisions <- data %>%
  select(State, Fatal_Collisions_Per_Billion, Percent_Speeding, Percent_Alcohol, Division) %>%
  pivot_longer(cols = c("Percent_Speeding", "Percent_Alcohol"),
               names_to = "Behavior",
               values_to = "Percentage")

```

To begin answering this question we grouped all of the states into 9 different U.S regional divisions. Then we reshaped the data to make it easier to plot.

```{r, fig.width=8, fig.height=10}
# Scatter plot of Fatal Collisions vs. Speeding, colored by Division
plot_speeding <- ggplot(speed_alcohol_collisions %>% filter(Behavior == "Percent_Speeding"),
                        aes(x = Percentage, y = Fatal_Collisions_Per_Billion, color = Division)) +
  geom_point(size = 3) + 
  geom_smooth(method = 'lm', se = FALSE) +
  labs(title = "Fatal Collisions vs. Percentage of Speeding Drivers",
       x = "Percentage of Speeding Drivers",
       y = "Fatal Collisions per Billion Miles",
       color = "US Division") +
  theme_minimal(base_size = 12) 

```


```{r, fig.width=8, fig.height=6}
# Scatter plot of Fatal Collisions vs. Alcohol Impairment, colored by Division
plot_alcohol <- ggplot(speed_alcohol_collisions %>% filter(Behavior == "Percent_Alcohol"),
                         aes(x = Percentage, y = Fatal_Collisions_Per_Billion, color = Division)) +
  geom_point(size = 3) + 
  geom_smooth(method = 'lm', se = FALSE) +
  labs(title = "Fatal Collisions vs. Percentage of Alcohol-Impaired Drivers",
       x = "Percentage of Alcohol-Impaired Drivers",
       y = "Fatal Collisions per Billion Miles",
       color = "US Division") +
  theme_minimal(base_size = 12)

grid.arrange(plot_speeding, plot_alcohol, ncol = 1)
```

The first plot we made shows the correlation between fatal collisions and drivers who were speeding while the second plot shows the correlation betweeen fatal collisions and drivers who were impaired by alcohol consumption. We decided to use a scatterplot with a line of best fit for each region because it displays the correlation for both speeding and alcohol columns to fatal collisions for each state and the average for the entire region. In the first plot there appears to be 6 positively correlated divisions and 3 negatively correlated divisions. For the second plot there are 5 positively correlated divisions and 4 negatively correlated divisions. Overall then we see that there is more of correlation between more fatal collisions due to alcohol and speeding throughout all the regions in the United States as we expected.
We did also find some regional patterns. 1, there is no strong correlation between alcohol use and more fatal collisions on the East coast. 2, there is no strong correlation between speeding and more fatal collisions in Western United states. Conversely, there is a strong correlation between alcohol and fatal collisions in Western United States. There is also a strong correlation between speeding and fatal collisions on the coasts of the United States.
So we can conclusively say that alcohol and speeding lead to more collisions and regional patterns do exist throughout the United States but the reason to why this is is beyond the scope of our project.

### Are there states where the percentage of speeding-related fatal collisions significantly deviates from the           national average, and what state-specific factors (e.g., speed limits, road conditions) might explain this? 

### How well can a predictive model (e.g., linear regression) use percentages of speeding, alcohol impairment, and       distraction to estimate fatal collisions per billion miles, and which factor is most influential?


## Conclusion

