---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis on Bad Drivers in the United States
#### Eitan Tuchin, Tirmidi Mohamed

## Introduction
The goal of our project is to explore data about bad drivers in the United States so that we can better understand causes of fatal car crashes and point out common characteristics about bad drivers. Since driving is such a common activity that we as humans do on a day-to-day basis it becomes imperative to analyze times at when driving has gone wrong in the name of trying to reduce the risk for these dangerous situations. We hope that our exploration and the analysis of the data we looked at will lead to more effective driving policies throughout America and an altogether smaller fatalities reported in later data.

To achieve our project's goal we will answer the following questions:

  1. Can we identify regional patterns in bad driving behaviors like speeding and alcohol impairment, and how do          these align with fatal collision rates? 

  2. Are there states where the percentage of speeding-related fatal collisions significantly deviates from the           national average, and what state-specific factors might explain this? 

  3. How well can a predictive model use percentages of speeding, alcohol impairment, and distraction to                  estimate fatal collisions per billion miles, and which factor is most influential?
  
These will be the main questions that we try and answer throughout our work on this project. We hope to get to some meaningful conclusions about bad drivers and their common trends in the United States.

## Data
### Structure

The link to our dataset is https://github.com/fivethirtyeight/data/blob/master/bad-drivers/bad-drivers.csv. The github link points to a csv file under a repository. Our dataset contains varying statistics relating to specifically bad-drivers throughout the United States including the District of Columbia. For each state (row) there are 7 columns that contain statistics like percentage of drivers involved in fatal collisions who were impaired by alcohol use. Each metric is numerical, either being an int or a float. Most deal with the "number of" or "percentage of" drivers in that state.

Overall our dataset is pretty simple and it contains all of the data we need for answering our questions. 

### Cleaning

The first thing we need to do is read in the csv file and take a look at it:

```{r}
data <- read.csv("bad-drivers.csv")
summary(data)
head(data)
```

The first thing we did cleaning wise was check for missing values in our data by iterating over every column in our data and checking if the value was NA or not. Then we removed any columns with missing values.

```{r}
missing_values <- colSums(is.na(data))
print("Missing values per column:")
print(missing_values)
data <- data[, colSums(is.na(data)) == 0]  
print("Columns after removing those with NAs:")
print(names(data))
```

Then, for simplicity sake, we wanted to rename some of the columns in our data because they weren't as data explorable and accessible friendly. For example, we have a column named "Number of drivers involved in fatal collisions per billion miles" which would be annoying to access since it's separated by white-space and very long. We also checked to see if we got the intended result.

```{r}
colnames(data) <- c("State", "Fatal_Collisions_Per_Billion", "Percent_Speeding", 
                    "Percent_Alcohol", "Percent_Not_Distracted", "Percent_No_Prior", 
                    "Insurance_Premiums", "Losses_Per_Driver")
str(data)
```

Since we are going to be doing some statistical modeling later to answer our question we decided to turn the State column into a factor type from a String. Then we grabbed all the columns that contained numeric data and stored them. We then ensured that these specified columns are treated as numbers and not anything else. Now we have easy data to work with.

```{r}
data$State <- as.factor(data$State)
numeric_cols <- c("Fatal_Collisions_Per_Billion", "Percent_Speeding", "Percent_Alcohol", 
                  "Percent_Not_Distracted", "Percent_No_Prior", "Insurance_Premiums", 
                  "Losses_Per_Driver")
data[numeric_cols] <- lapply(data[numeric_cols], as.numeric)
```

Lastly, we need to check if we have any duplicate rows in our data. Then we removed those rows from the data.

```{r}
duplicates <- duplicated(data)
print("Number of duplicate rows:")
print(sum(duplicates))

data <- data[!duplicates, ]  

print("Final cleaned data:")
head(data)
```

After cleaning our data it's ready to be analyzed and interpreted.

### Variables

Here we list all of the variables (columns) in our dataset:

* State - The state to which the statistics in the row are attributed to.
* Fatal_Collisions_Per_Billion - Number of drivers involved in fatal collisions per billion miles.
* Percent_Speeding - Percentage of drivers who were involved in fatal collisions who were speeding.
* Percent_Alcohol - Percentage of drivers who were involved in fatal collisions who were alcohol-impaired.
* Percent_Not_Distracted - Percentage of drivers who were involved in fatal collisions who were not distracted.
* Percent_No_Prior - Percentage of drivers who were involved in fatal collisions who had not been in any previous accidents
* Insurance_Premiums - Average cost of car insurance premiums in that state.
* Losses_Per_Driver - Losses incurred by insurance companies for collisions per insured driver.

## Results

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(reshape2)
library(gridExtra)
```


### Can we identify regional patterns in bad driving behaviors like speeding and alcohol impairment, and how do these     align with fatal collision rates? 

```{r}
divisions <- list(
  `New England` = c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont"),
  `Middle Atlantic` = c("New Jersey", "New York", "Pennsylvania"),
  `East North Central` = c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin"),
  `West North Central` = c("Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska", "North Dakota", "South Dakota"),
  `South Atlantic` = c("Delaware", "Florida", "Georgia", "Maryland", "North Carolina", "South Carolina", 
                      "Virginia", "West Virginia", "District of Columbia"),
  `East South Central` = c("Alabama", "Kentucky", "Mississippi", "Tennessee"),
  `West South Central` = c("Arkansas", "Louisiana", "Oklahoma", "Texas"),
  `Mountain` = c("Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Utah", "Wyoming"),
  `Pacific` = c("Alaska", "California", "Hawaii", "Oregon", "Washington")
)

data <- data %>%
  mutate(Division = case_when(
    State %in% divisions$`New England` ~ "New England",
    State %in% divisions$`Middle Atlantic` ~ "Middle Atlantic",
    State %in% divisions$`East North Central` ~ "East North Central",
    State %in% divisions$`West North Central` ~ "West North Central",
    State %in% divisions$`South Atlantic` ~ "South Atlantic",
    State %in% divisions$`East South Central` ~ "East South Central",
    State %in% divisions$`West South Central` ~ "West South Central",
    State %in% divisions$`Mountain` ~ "Mountain",
    State %in% divisions$`Pacific` ~ "Pacific",
    TRUE ~ NA_character_
  ))


speed_alcohol_collisions <- data %>%
  select(State, Fatal_Collisions_Per_Billion, Percent_Speeding, Percent_Alcohol, Division) %>%
  pivot_longer(cols = c("Percent_Speeding", "Percent_Alcohol"),
               names_to = "Behavior",
               values_to = "Percentage")

```

To begin answering this question we grouped all of the states into 9 different U.S regional divisions. Then we reshaped the data to make it easier to plot.

```{r, fig.width=8, fig.height=10}
# Scatter plot of Fatal Collisions vs. Speeding, colored by Division
plot_speeding <- ggplot(speed_alcohol_collisions %>% filter(Behavior == "Percent_Speeding"),
                        aes(x = Percentage, y = Fatal_Collisions_Per_Billion, color = Division)) +
  geom_point(size = 3) + 
  geom_smooth(method = 'lm', se = FALSE) +
  labs(title = "Fatal Collisions vs. Percentage of Speeding Drivers",
       x = "Percentage of Speeding Drivers",
       y = "Fatal Collisions per Billion Miles",
       color = "US Division") +
  theme_minimal(base_size = 12) 

```


```{r, fig.width=8, fig.height=6}
# Scatter plot of Fatal Collisions vs. Alcohol Impairment, colored by Division
plot_alcohol <- ggplot(speed_alcohol_collisions %>% filter(Behavior == "Percent_Alcohol"),
                         aes(x = Percentage, y = Fatal_Collisions_Per_Billion, color = Division)) +
  geom_point(size = 3) + 
  geom_smooth(method = 'lm', se = FALSE) +
  labs(title = "Fatal Collisions vs. Percentage of Alcohol-Impaired Drivers",
       x = "Percentage of Alcohol-Impaired Drivers",
       y = "Fatal Collisions per Billion Miles",
       color = "US Division") +
  theme_minimal(base_size = 12)

grid.arrange(plot_speeding, plot_alcohol, ncol = 1)
```

The first plot we made shows the correlation between fatal collisions and drivers who were speeding while the second plot shows the correlation betweeen fatal collisions and drivers who were impaired by alcohol consumption. We decided to use a scatterplot with a line of best fit for each region because it displays the correlation for both speeding and alcohol columns to fatal collisions for each state and the average for the entire region. In the first plot there appears to be 6 positively correlated divisions and 3 negatively correlated divisions. For the second plot there are 5 positively correlated divisions and 4 negatively correlated divisions. Overall then we see that there is more of correlation between more fatal collisions due to alcohol and speeding throughout all the regions in the United States as we expected.
We did also find some regional patterns. 1, there is no strong correlation between alcohol use and more fatal collisions on the East coast. 2, there is no strong correlation between speeding and more fatal collisions in Western United states. Conversely, there is a strong correlation between alcohol and fatal collisions in Western United States. There is also a strong correlation between speeding and fatal collisions on the coasts of the United States.
So we can conclusively say that alcohol and speeding lead to more collisions and regional patterns do exist throughout the United States but the reason to why this is is beyond the scope of our project.

### Are there states where the percentage of speeding-related fatal collisions significantly deviates from the           national average, and what state-specific factors (e.g., speed limits, road conditions) might explain this? 

To explore states with unusual speeding-related fatality percentages, we first calculated the national average and standard deviation for the `Percent_Speeding` column. We used these values to determine how many standard deviations each state's percentage lies away from the national mean. This helps us identify states that are statistical outliers.

```{r speeding-deviation-calculation}
# Calculate national average and standard deviation for speeding percentage
national_avg_speeding <- mean(data$Percent_Speeding, na.rm = TRUE)
national_sd_speeding <- sd(data$Percent_Speeding, na.rm = TRUE)

# Print the national average and SD
print(paste("National Average Percentage Speeding:", round(national_avg_speeding, 2)))
print(paste("National Standard Deviation Percentage Speeding:", round(national_sd_speeding, 2)))

# Add deviation in SD units to the data frame
data <- data %>%
  mutate(Speeding_Deviation_SD = (Percent_Speeding - national_avg_speeding) / national_sd_speeding)

# Identify states deviating significantly (e.g., more than 1.5 SD from the mean)
significant_deviators <- data %>%
  filter(abs(Speeding_Deviation_SD) > 1.5) %>%
  
  arrange(desc(Speeding_Deviation_SD))

print("States with speeding percentages > 1.5 SD from national average:")
print(significant_deviators %>% select(State, Percent_Speeding, Speeding_Deviation_SD))
```
We then visualize the percentage of speeding-related fatal collisions for all states, highlighting the national average. This allows for easy comparison and identification of states at the extremes.
```{r}
# Plot Percentage Speeding by State, ordered, with national average line
plot_speeding_states <- ggplot(data, aes(x = reorder(State, Percent_Speeding), y = Percent_Speeding)) +
  geom_col(aes(fill = Speeding_Deviation_SD > 1.5 | Speeding_Deviation_SD < -1.5)) + # Highlight significant deviators
  geom_hline(yintercept = national_avg_speeding, color = "red", linetype = "dashed") +
  scale_fill_manual(values = c("grey", "orange"), name = "Deviation > 1.5 SD") +
  coord_flip() + # Flip coordinates for better state label readability
  labs(title = "Percentage of Speeding-Related Fatal Collisions by State",
       x = "State",
       y = "Percentage of Drivers Speeding in Fatal Collisions",
       caption = "Red dashed line indicates the national average.") +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_text(size = 8)) # Adjust y-axis text size if needed

print(plot_speeding_states)
```

The plot reveals considerable variation across states. States like Hawaii and Pennsylvania stand out with significantly higher percentages of speeding-related fatal collisions compared to the national average (indicated in orange). Conversely, states like Mississippi and Nebraska show significantly lower percentages.

While our dataset doesn't include direct state-specific factors to definitively explain these deviations, potential contributing factors could include:
* **Speed Limits:** States with higher maximum speed limits might see more speeding, although the relationship with *fatal* speeding crashes could be complex.
* **Road Infrastructure:** Differences in highway quality, urban vs. rural road mileage, and topography.
* **Traffic Enforcement:** Varying levels of speed limit enforcement and penalties.
* **Driving Culture & Demographics:** Differences in local driving norms, population density, and age distribution of drivers.
* **Weather Conditions:** States with more adverse weather might see different patterns, potentially reducing speeds overall but increasing risk when speeding does occur.

Further analysis incorporating these external factors would be needed for a more conclusive explanation.
### How well can a predictive model (e.g., linear regression) use percentages of speeding, alcohol impairment, and       distraction to estimate fatal collisions per billion miles, and which factor is most influential?

To assess the predictive power of key behavioral factors on fatal collision rates, we constructed a multiple linear regression model. The model aims to predict `Fatal_Collisions_Per_Billion` using `Percent_Speeding`, `Percent_Alcohol`, and `Percent_Not_Distracted` as predictor variables. It's important to remember that `Percent_Not_Distracted` represents drivers who were *not* distracted; a higher value implies less distraction was involved in fatal crashes, so we might expect it to have a negative relationship with the overall fatal collision rate.

```{r linear-model}
# Build the linear regression model using the cleaned data
collision_model <- lm(Fatal_Collisions_Per_Billion ~ Percent_Speeding + Percent_Alcohol + Percent_Not_Distracted, data = data)

# Display the model summary which contains coefficients, p-values, R-squared, etc.
summary(collision_model)
```

**Interpretation of the Model:**

The model summary provides insights into how well these factors collectively predict fatal collisions and the individual influence of each factor.

* **Model Fit ("How well?"):** The R-squared value tells us the proportion of variance in `Fatal_Collisions_Per_Billion` that is explained by the three predictors combined. The **Adjusted R-squared value is -0.01275**. An adjusted R-squared this low (essentially zero, and even slightly negative) indicates that the model has **no explanatory power**. It means that these three factors, combined in this linear model, do not explain the variation in fatal collisions per billion miles across states any better than simply guessing the average rate. The overall F-statistic p-value of 0.5055 further confirms that the model as a whole is not statistically significant.

* **Factor Influence ("Which factor is most influential?"):** We examine the coefficients and their corresponding p-values (the `Estimate` and `Pr(>|t|)` columns in the summary table) to determine the significance and influence of each predictor:
    * `Percent_Speeding`: The coefficient is -0.040857 with a p-value of 0.5267. Based on the p-value (which is much greater than the standard 0.05 threshold), this factor is **not statistically significant** in predicting fatal collisions within this model.
    * `Percent_Alcohol`: The coefficient is 0.181687 with a p-value of 0.1345. Based on the p-value (which is greater than 0.05), this factor is also **not statistically significant**. While the coefficient is positive (suggesting a potential positive association), the evidence is not strong enough to conclude a reliable relationship exists in this dataset.
    * `Percent_Not_Distracted`: The coefficient is 0.003412 with a p-value of 0.9307. This factor is also **not statistically significant**. The coefficient is very close to zero and the p-value is very high, indicating no detectable relationship between the percentage of non-distracted drivers in fatal crashes and the overall fatal collision rate in this model.

* **Most Influential Factor:** Since **none** of the predictors (`Percent_Speeding`, `Percent_Alcohol`, `Percent_Not_Distracted`) were found to be statistically significant (all p-values > 0.05), **we cannot identify a 'most influential' factor based on this linear model**. The model fails to establish a statistically reliable link between these specific behavioral percentages and the rate of fatal collisions per billion miles across states.

In summary, the linear model attempted here performs very poorly (Adjusted R-squared ≈ -0.013) and does not provide statistically significant evidence that the state-level percentages of speeding, alcohol impairment, or non-distraction (as measured in this dataset) are reliable predictors of the overall fatal collision rate per billion miles. There might be more complex relationships, issues with the data aggregation at the state level, or other unmeasured factors that are more important.

```{r}
# 📈 Graph 1: Fatal Collisions vs Percent Alcohol (Simple Scatter)
ggplot(data, aes(x = Percent_Alcohol, y = Fatal_Collisions_Per_Billion)) +
  geom_point(color = "blue", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Fatal Collisions vs Percentage of Alcohol-Impaired Drivers",
       x = "Percentage of Alcohol-Impaired Drivers",
       y = "Fatal Collisions per Billion Miles") +
  theme_minimal()



# 📈 Graph 2: Predicted vs Actual Fatal Collisions (Smoothed)
# Predict values again if needed
data$Predicted_Collisions <- predict(collision_model)

ggplot(data, aes(x = Fatal_Collisions_Per_Billion, y = Predicted_Collisions)) +
  geom_point(color = "darkorange", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "steelblue") +
  labs(title = "Predicted vs Actual Fatal Collisions",
       x = "Actual Fatal Collisions per Billion Miles",
       y = "Predicted Fatal Collisions per Billion Miles") +
  theme_minimal()

```

## Conclusion

Our goal in this project was to use state-level data to get a clearer picture of bad driving behaviors in the United States, specifically looking at speeding and alcohol impairment and how they connect to fatal crash rates. We explored regional differences, looked for states that stood out, and tested whether common factors could predict fatal collisions per billion miles. While our analysis provided some useful insights, it also clearly showed the challenges of working with state-level aggregated data for a complex issue like traffic safety.

Looking at regional patterns (Question 1), we did see that, generally, higher rates of speeding and alcohol impairment tend to go along with higher fatal collision rates, as one might expect. However, this wasn't consistent across all regions. We noticed distinct variations—for example, the link between alcohol and fatalities seemed less pronounced on the East Coast compared to the West. This suggests that regional context matters, and factors not included in our dataset likely play a significant role.

In Question 2, we successfully pinpointed states where the percentage of speeding drivers in fatal crashes was significantly different from the national average. Hawaii and Pennsylvania, for instance, were notably higher, while Mississippi and Nebraska were lower. This confirms that driving safety issues related to speeding vary considerably from state to state. Although we could speculate about reasons like different speed limits, road types, or enforcement levels, our data didn't allow us to definitively explain why these states stand out.

The results from Question 3 were particularly revealing. Our attempt to predict fatal collisions per billion miles using a standard linear regression model with state percentages for speeding, alcohol impairment, and non-distraction was unsuccessful. The model had virtually no predictive power (Adjusted R² ≈ -0.013), and critically, none of these factors turned out to be statistically significant predictors of the fatal collision rate at the state level. This important finding doesn't mean these behaviors aren't dangerous, but rather that using these specific state-wide average percentages in a simple linear model isn't enough to reliably predict the overall fatal collision rate across states.

Taken together, our findings illustrate that while state-level data can reveal broad patterns and identify outliers, predicting fatal crash rates is complex. The factors involved likely interact in ways that aren't captured well by state averages or simple linear relationships. Key limitations we faced include the nature of aggregated data (state averages might hide important local variations), the absence of data on crucial factors like traffic enforcement intensity, specific road conditions, or detailed driver information, and the constraints of the linear model we applied.

For future work, exploring this topic with more granular data—perhaps at the county level or using individual crash records—could be more fruitful. Incorporating a wider array of variables and potentially employing more advanced modeling techniques might also uncover relationships missed here. Despite the predictive limitations found in our analysis, the variations observed across states and regions reinforce the ongoing importance of addressing dangerous driving behaviors like speeding and alcohol impairment through targeted safety efforts nationwide.
