---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis on Fatal Motor Vehicle Accidents in the United States
#### Marissa Baietto, Ryan Dorle, Braeden Collings

## Introduction
The goal of the project is to explore multiple datasets to better understand fatal vehicle crashes. Unfortunately, fatal car crashes have become increasingly prevalent due to the high volume of drivers in the United States. Analysis on this topic can lead to increased awareness, more effective policing efforts, and identification of common trends for accidents. Ultimately, our goal is to understand the circumstances in which individuals get into vehicular accidents.

In pursuit of the stated goal, we will explore the following questions:

1. When are crashes most likely? Are there any seasonal effects, and are night crashes more likely than in the morning or afternoon?

2. How is speeding related to the number of fatalities? Are younger drivers more prone to speeding? Is higher speeding limits associated with more fatalities?

3. Does impairment affect fatality in crashes overall? When are impaired crashes most likely?

4. What regions of the United States have the most fatal crashes? What conditions are present in those regions?

5. How does the demographics of the driver affect crashes? Are changes based on occupants more prevalent for younger drivers?

6. Are crashes affected by lighting and road conditions? How so, and what conditions are most impact? 

These are the main questions we are looking to answer through the completion of this project. With the findings we will be able to draw meaningful conclusions on fatal car crashes throughout the United States. 

## Data

## Results
```{r, warning=FALSE,message=FALSE}
library(readr)
library(tidyverse)
```

## Conclusion

